---
title: "S&P 500 Financial Analytics"
subtitle: "Profitability, Valuation, Risk & Predictive Modelling of 496 U.S. Companies"
author: "Abdulrahman Jalilov"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      smooth_scroll: true
    toc_depth: 3
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 10, fig.height = 6)

# Brand colors (consistent with Retail & HR projects)
COL_TEAL   <- "#2A9D8F"
COL_RED    <- "#E76F51"
COL_NAVY   <- "#2C3E50"
COL_BLUE   <- "#457B9D"
COL_ORANGE <- "#F4A261"
COL_PURPLE <- "#4B4B8F"
```

---

# Executive Summary

This report analyses **496 S&P 500 companies** across 74 financial indicators sourced from Yahoo Finance / Finviz, stored in a MySQL database (`sp500_project.companies`), and exported via a structured SQL pipeline of 17 queries across 8 analytical sections. **The data was captured in early 2023** — during the bear market recovery phase following the 2022 S&P 500 decline of ~19% — which provides critical context for interpreting the forecasting results.

**Business Problem:** The S&P 500 is widely followed as a benchmark, but individual investors and analysts face a critical challenge — identifying which companies offer the best combination of profitability, financial stability, valuation, and growth potential. This project answers three questions:

1. **Where is value being created?** (Profitability Analysis)
2. **Where is risk being taken?** (Valuation & Stability)
3. **What drives short-term stock performance?** (Regression Forecasting)

**Key Findings:**

- Momentum (RSI) explains approximately **56% of YTD performance** — far more than any fundamental metric
- Fundamentals alone (9 ratios) explain only approximately **16%** of short-term price variation
- The "Financial Fortress" screen identified only **15 companies** (3% of the index) passing all four quality criteria simultaneously
- Three failed banks (SVB, Signature, First Republic) appeared as "undervalued" in the data — the most important data quality warning in the project: **no financial ratio captures bank run risk**

**Tools:** MySQL (17 queries, 8 analytical sections) → R (cleaning, EDA, regression) → Excel/Shiny (dashboard)

---

# 1. Data Loading & Connection

## 1.1 MySQL Connection

This report connects to a local MySQL database built from the DataCo S&P 500 dataset. The connection uses `DBI` + `RMariaDB`. For the deployed report, we load the pre-exported CSV from Query 9.1.

```{r mysql_connection, eval=FALSE}
# Live MySQL connection (run locally — not executed in knit)
library(DBI)
library(RMariaDB)

con <- dbConnect(
  RMariaDB::MariaDB(),
  host     = "127.0.0.1",
  port     = 3306,
  dbname   = "sp500_project",
  user     = "root",
  password = rstudioapi::askForPassword("MySQL password")
)

# This produces the sp500_clean_export.csv used below
df_raw_live <- dbGetQuery(con, "
  SELECT ticker, index_membership,
    CAST(REPLACE(pe_ratio,'-','') AS DECIMAL(10,2)) AS pe_ratio,
    CAST(REPLACE(net_profit_margin,'%','') AS DECIMAL(10,2)) AS net_margin,
    CAST(REPLACE(beta,'-','') AS DECIMAL(10,2)) AS beta,
    CAST(REPLACE(perf_ytd,'%','') AS DECIMAL(10,2)) AS ytd_performance,
    CAST(REPLACE(rsi,'-','') AS DECIMAL(10,2)) AS rsi
  FROM companies
  WHERE pe_ratio NOT LIKE '-%' AND pe_ratio != '-'
    AND net_profit_margin != '-' AND beta != '-'
    AND perf_ytd != '-' AND rsi != '-';
")

dbDisconnect(con)
```

## 1.2 Load Exported Data

```{r load_data}
library(tidyverse)
library(caret)
library(corrplot)
library(kableExtra)
library(scales)
library(broom)

# Load from MySQL Query 9.1 export
# File: sp500_clean_export.csv
# Place this file in the same directory as this .Rmd
df_raw <- read.csv("sp500_clean_export.csv", stringsAsFactors = FALSE)

cat("Raw dataset:", nrow(df_raw), "companies,", ncol(df_raw), "variables\n")
cat("YTD Performance: ", round(min(df_raw$ytd_performance), 2),
    "% to", round(max(df_raw$ytd_performance), 2), "%\n")
```

## 1.3 Data Cleaning

```{r cleaning}
# Remove statistical outliers that would distort regression coefficients:
# - Debt-inflated ROE (APA = 924.9%, MSI = -518.7%) → cap at ±200%
# - Distorted P/E (CRM = 888x, GE = 160x restructuring) → remove > 200
# - Extreme leverage (NCLH D/E = 198.56) → remove > 20
# - Loss-making companies → net_margin >= 0

df_clean <- df_raw %>%
  filter(
    roe          >  -200  & roe          < 200,
    pe_ratio     <   200,
    debt_to_equity <  20,
    net_margin   >=    0
  ) %>%
  drop_na(rsi, beta, ytd_performance, net_margin, roe, pe_ratio)

removed <- nrow(df_raw) - nrow(df_clean)

summary_table <- data.frame(
  Stage       = c("Raw Export (Query 9.1)", "After Outlier Removal"),
  Companies   = c(nrow(df_raw), nrow(df_clean)),
  Notes       = c(
    "Includes debt-inflated ROE, distorted P/E, extreme leverage",
    paste0(removed, " outliers removed (", round(removed/nrow(df_raw)*100,1), "%)")
  )
)

summary_table %>%
  kable(caption = "Data Cleaning Summary") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped","hover","condensed","responsive"),
                position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = COL_PURPLE)
```

> **Note:** Outlier companies removed include APA (ROE 924.9%), NCLH (D/E 198.56), CRM (P/E 888), and loss-making companies. SVB, Signature Bank, and First Republic were also removed as they subsequently failed — their financial ratios gave no warning of bank run risk, which is the most important data quality caveat of this project.

---

# 2. Descriptive Statistics

## 2.1 YTD Performance Overview

```{r ytd_overview}
ytd_stats <- data.frame(
  Metric = c("Mean", "Median", "Std Dev", "Min", "Max",
             "% Positive", "% Negative"),
  Value = c(
    paste0(round(mean(df_clean$ytd_performance), 2), "%"),
    paste0(round(median(df_clean$ytd_performance), 2), "%"),
    paste0(round(sd(df_clean$ytd_performance), 2), "%"),
    paste0(round(min(df_clean$ytd_performance), 2), "%"),
    paste0(round(max(df_clean$ytd_performance), 2), "%"),
    paste0(round(mean(df_clean$ytd_performance > 0) * 100, 1), "%"),
    paste0(round(mean(df_clean$ytd_performance < 0) * 100, 1), "%")
  )
)

ytd_stats %>%
  kable(caption = "YTD Performance Summary (After Cleaning, 427 Companies)") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped","hover"),
                position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = COL_PURPLE)
```

## 2.2 YTD Performance Distribution

```{r ytd_distribution}
ggplot(df_clean, aes(x = ytd_performance)) +
  geom_histogram(bins = 40, fill = COL_BLUE, color = "white", alpha = 0.85) +
  geom_vline(xintercept = 0,
             color = COL_RED, linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = mean(df_clean$ytd_performance),
             color = COL_TEAL, linewidth = 1.2) +
  annotate("text", x = mean(df_clean$ytd_performance) + 3,
           y = 30, label = paste0("Mean: ",
           round(mean(df_clean$ytd_performance), 1), "%"),
           color = COL_TEAL, fontface = "bold", size = 4) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(
    title    = "Distribution of YTD Stock Performance — S&P 500 (Early 2023)",
    subtitle = "Data captured during the bear market recovery phase",
    x        = "YTD Performance (%)",
    y        = "Number of Companies",
    caption  = "Red dashed = 0% breakeven | Teal = mean return"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 16, color = COL_NAVY),
    plot.subtitle = element_text(color = "gray40"),
    plot.caption  = element_text(color = "gray40"),
    panel.grid.minor = element_blank()
  )
```

> **Insight:** The distribution is roughly normal with a slight positive skew — the mean (+3%) is above the median (+1.7%), indicating a small number of extreme outperformers (NVDA +64%, META +53%) are pulling the average up. Data was captured during the early 2023 recovery after the 2022 bear market, which explains why the majority (293 of 427 companies) still had RSI in bearish territory (30-50).

## 2.3 Key Financial Metrics by Index Membership

```{r metrics_by_index}
index_summary <- df_clean %>%
  group_by(index_membership) %>%
  summarise(
    n             = n(),
    avg_ytd       = round(mean(ytd_performance, na.rm = TRUE), 2),
    avg_pe        = round(mean(pe_ratio, na.rm = TRUE), 2),
    avg_net_margin= round(mean(net_margin, na.rm = TRUE), 2),
    avg_roe       = round(mean(roe, na.rm = TRUE), 2),
    avg_beta      = round(mean(beta, na.rm = TRUE), 2),
    avg_rsi       = round(mean(rsi, na.rm = TRUE), 2),
    .groups       = "drop"
  ) %>%
  arrange(desc(avg_net_margin)) %>%
  rename(
    `Index`        = index_membership,
    `Companies`    = n,
    `Avg YTD %`    = avg_ytd,
    `Avg P/E`      = avg_pe,
    `Avg Net Margin%` = avg_net_margin,
    `Avg ROE%`     = avg_roe,
    `Avg Beta`     = avg_beta,
    `Avg RSI`      = avg_rsi
  )

index_summary %>%
  kable(caption = "Financial Metrics by Index Membership (Cleaned Dataset)") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped","hover","condensed"),
                position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = COL_PURPLE)
```

> **Insight:** NDX + S&P 500 companies lead on net margin — consistent with the MySQL analysis showing 57.94% average gross margin for NASDAQ-100 companies. The 7 "triple-index" companies (DJIA + NDX + S&P 500) show the lowest average Beta (0.92 — most defensive) yet among the highest YTD performance, confirming that quality and stability are rewarded even in recovery markets.

---

# 3. Correlation Analysis

## 3.1 Correlation Heatmap

```{r correlation_heatmap, fig.width=11, fig.height=9}
numeric_df <- df_clean %>%
  select(ytd_performance, pe_ratio, pb_ratio, net_margin, gross_margin,
         operating_margin, roe, roa, debt_to_equity, current_ratio,
         beta, eps_growth_fwd, lt_growth_5y, dividend_yield, rsi, analyst_rec)

cor_matrix <- cor(numeric_df, use = "complete.obs")

corrplot(cor_matrix,
         method   = "color",
         type     = "upper",
         tl.cex   = 0.85,
         tl.col   = COL_NAVY,
         addCoef.col = "black",
         number.cex  = 0.6,
         col      = colorRampPalette(c(COL_RED, "white", COL_TEAL))(200),
         mar      = c(0, 0, 2, 0),
         title    = "Correlation Matrix — S&P 500 Financial Indicators")
```

## 3.2 Top Correlations with YTD Performance

```{r top_correlations}
corr_with_ytd <- cor_matrix["ytd_performance", ] %>%
  as.data.frame() %>%
  rownames_to_column("Variable") %>%
  rename(Correlation = ".") %>%
  filter(Variable != "ytd_performance") %>%
  mutate(
    Direction = ifelse(Correlation > 0, "Positive", "Negative"),
    Correlation = round(Correlation, 3)
  ) %>%
  arrange(desc(abs(Correlation)))

corr_with_ytd %>%
  kable(caption = "Correlations with YTD Performance — Ranked by Absolute Value") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped","hover"),
                position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = COL_PURPLE) %>%
  row_spec(1, bold = TRUE, background = "#d4edda") %>%
  row_spec(which(corr_with_ytd$Correlation < -0.2) + 1,
           background = "#fde8e8")
```

> **Insight:** RSI has by far the strongest correlation with YTD performance (r = **0.69**) — higher than any fundamental metric. This is the central finding of the analysis: in the short run, momentum (price strength) explains far more of stock performance than financial quality. Beta (r = 0.30) ranks second — higher-risk stocks recovered more aggressively in the early 2023 rebound. Dividend yield shows the strongest negative relationship (r = -0.30) — high-yield defensive stocks lagged the recovery, consistent with the sector rotation we observed in the MySQL dividend analysis.

> **⚠️ Methodological Note:** RSI is calculated from prior price movements, which already contain partial YTD return information. The 0.69 correlation therefore reflects both genuine momentum effects *and* some degree of circular measurement. The causal interpretation should be treated with appropriate caution — RSI identifies *which stocks were already rising*, not necessarily *why* they would continue to rise. This is why academic momentum research (Jegadeesh & Titman, 1993) carefully tests momentum on *future* returns over 3-12 month windows rather than contemporaneous YTD snapshots.

---

# 4. Regression Modelling

## 4.1 Model Strategy

We build three models of increasing complexity to test the core question: **do financial fundamentals predict YTD performance, or does momentum dominate?**

| Model | Variables | Hypothesis |
|---|---|---|
| **Model A — Momentum** | RSI, Beta, Dividend Yield | Short-term performance is driven by price momentum |
| **Model B — Fundamentals** | 9 financial ratios | Financial quality determines stock performance |
| **Model C — Full** | All 13 variables | Combined model captures both effects |

## 4.2 Data Preparation

```{r model_prep}
set.seed(42)

features_momentum    <- c("rsi", "beta", "dividend_yield")
features_fundamental <- c("pe_ratio", "net_margin", "roe", "roa",
                           "debt_to_equity", "current_ratio",
                           "gross_margin", "eps_growth_fwd", "lt_growth_5y")
features_full        <- c(features_momentum, features_fundamental, "analyst_rec")

# Scale features for coefficient comparability
df_model <- df_clean %>%
  select(ticker, ytd_performance, all_of(features_full)) %>%
  drop_na()

# Standardize predictors
preproc  <- preProcess(df_model[, features_full], method = c("center", "scale"))
df_scaled <- predict(preproc, df_model)

# 80/20 train/test split
train_idx  <- createDataPartition(df_scaled$ytd_performance, p = 0.8, list = FALSE)
train_data <- df_scaled[train_idx, ]
test_data  <- df_scaled[-train_idx, ]

cat("Training set:", nrow(train_data), "companies\n")
cat("Test set:    ", nrow(test_data), "companies\n")
```

## 4.3 Model A — Momentum Only

```{r model_momentum}
model_a <- lm(ytd_performance ~ rsi + beta + dividend_yield,
              data = train_data)
summary(model_a)
```

## 4.4 Model B — Fundamentals Only

```{r model_fundamentals}
model_b <- lm(
  ytd_performance ~ pe_ratio + net_margin + roe + roa +
    debt_to_equity + current_ratio + gross_margin +
    eps_growth_fwd + lt_growth_5y,
  data = train_data
)
summary(model_b)
```

## 4.5 Model C — Full Model

```{r model_full}
model_c <- lm(
  ytd_performance ~ rsi + beta + dividend_yield +
    pe_ratio + net_margin + roe + roa +
    debt_to_equity + current_ratio + gross_margin +
    eps_growth_fwd + lt_growth_5y + analyst_rec,
  data = train_data
)
summary(model_c)
```

## 4.6 Model Comparison

```{r model_comparison}
ctrl <- trainControl(method = "cv", number = 5)

cv_a <- train(ytd_performance ~ rsi + beta + dividend_yield,
              data = train_data, method = "lm", trControl = ctrl)
cv_b <- train(ytd_performance ~ pe_ratio + net_margin + roe + roa +
                debt_to_equity + current_ratio + gross_margin +
                eps_growth_fwd + lt_growth_5y,
              data = train_data, method = "lm", trControl = ctrl)
cv_c <- train(ytd_performance ~ .,
              data = train_data %>% select(-ticker),
              method = "lm", trControl = ctrl)

# Test set performance
pred_a <- predict(model_a, newdata = test_data)
pred_b <- predict(model_b, newdata = test_data)
pred_c <- predict(model_c, newdata = test_data)

rmse_fn <- function(actual, predicted)
  sqrt(mean((actual - predicted)^2, na.rm = TRUE))

model_compare <- data.frame(
  Model        = c("A: Momentum Only", "B: Fundamentals Only", "C: Full Model"),
  Variables    = c(3, 9, 13),
  Train_R2     = c(round(summary(model_a)$r.squared, 4),
                   round(summary(model_b)$r.squared, 4),
                   round(summary(model_c)$r.squared, 4)),
  CV_R2        = c(round(cv_a$results$Rsquared, 4),
                   round(cv_b$results$Rsquared, 4),
                   round(cv_c$results$Rsquared, 4)),
  Test_R2      = c(round(cor(pred_a, test_data$ytd_performance)^2, 4),
                   round(cor(pred_b, test_data$ytd_performance, use="complete.obs")^2, 4),
                   round(cor(pred_c, test_data$ytd_performance)^2, 4)),
  Test_RMSE    = c(round(rmse_fn(test_data$ytd_performance, pred_a), 3),
                   round(rmse_fn(test_data$ytd_performance, pred_b), 3),
                   round(rmse_fn(test_data$ytd_performance, pred_c), 3))
)

model_compare %>%
  kable(caption = "Model Comparison — R², Cross-Validated R², Test RMSE") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped","hover","condensed"),
                position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = COL_PURPLE) %>%
  row_spec(1, bold = TRUE, background = "#d4edda") %>%
  row_spec(3, bold = TRUE, background = "#e8f4f8")
```

> **Insight:** Model A (Momentum Only) with just 3 variables achieves CV R² = **`r round(cv_a$results$Rsquared, 3)`** — explaining over half of YTD performance variation. Model B (Fundamentals, 9 variables) achieves only CV R² = **`r round(cv_b$results$Rsquared, 3)`** — fundamental ratios have minimal short-term predictive power. Adding fundamentals to momentum (Model C) improves R² only marginally — confirming that **momentum dominates short-term stock returns**, consistent with the academic "momentum factor" documented by Jegadeesh & Titman (1993) and widely used in quantitative finance.

> **Note on R²:** An R² of ~0.56 for cross-sectional stock return prediction is genuinely high by financial research standards. Academic literature typically reports R² of 0.05-0.30 for fundamental models and 0.10-0.40 for momentum models. The unusually high momentum R² here reflects the sharp, RSI-driven dynamics of the early 2023 recovery — a market regime where momentum was the dominant force.

## 4.7 Model R² Comparison Chart

```{r model_comparison_chart}
model_compare_long <- model_compare %>%
  select(Model, Train_R2, CV_R2, Test_R2) %>%
  pivot_longer(cols = c(Train_R2, CV_R2, Test_R2),
               names_to = "Metric", values_to = "R2") %>%
  mutate(
    Metric = recode(Metric,
                    "Train_R2" = "Train R²",
                    "CV_R2"    = "CV R² (5-fold)",
                    "Test_R2"  = "Test R²"),
    Metric = factor(Metric, levels = c("Train R²", "CV R² (5-fold)", "Test R²"))
  )

ggplot(model_compare_long,
       aes(x = Model, y = R2, fill = Metric)) +
  geom_col(position = "dodge", width = 0.65) +
  geom_text(aes(label = round(R2, 3)),
            position = position_dodge(width = 0.65),
            vjust = -0.4, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c(COL_NAVY, COL_TEAL, COL_BLUE)) +
  scale_y_continuous(limits = c(0, 0.70), labels = number_format(accuracy = 0.01)) +
  geom_hline(yintercept = 0.3, linetype = "dashed",
             color = "gray50", linewidth = 0.8) +
  annotate("text", x = 0.6, y = 0.31,
           label = "Academic benchmark (~0.30)", color = "gray50", size = 3.5) +
  labs(
    title    = "Model R² Comparison — Momentum vs Fundamentals vs Full",
    subtitle = "Momentum (3 variables) outperforms Fundamentals (9 variables) by a wide margin",
    x        = NULL,
    y        = "R² (proportion of variance explained)",
    fill     = "Metric",
    caption  = "Dashed line = typical academic benchmark for cross-sectional return models"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 16, color = COL_NAVY),
    plot.subtitle = element_text(color = "gray40"),
    plot.caption  = element_text(color = "gray40"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

> **Insight:** The gap between the Momentum and Fundamentals bars is the visual core of this analysis. Despite having 3x as many variables, the Fundamentals model achieves less than one-third of the momentum model's explanatory power. The Full Model barely improves over Momentum alone — adding 10 more fundamental variables contributes almost nothing. This is not a failure of the fundamentals model — it is the expected result and confirms 30+ years of empirical asset pricing research.

## 4.8 Top & Bottom Performers — The RSI Pattern

```{r top_bottom_performers}
# Top 10 by YTD performance
top10 <- df_clean %>%
  arrange(desc(ytd_performance)) %>%
  slice(1:10) %>%
  select(ticker, ytd_performance, rsi, beta, net_margin) %>%
  mutate(Group = "Top 10")

# Bottom 10 by YTD performance
bot10 <- df_clean %>%
  arrange(ytd_performance) %>%
  slice(1:10) %>%
  select(ticker, ytd_performance, rsi, beta, net_margin) %>%
  mutate(Group = "Bottom 10")

performers <- bind_rows(top10, bot10)

# Table
performers %>%
  select(-Group) %>%
  mutate(
    ytd_performance = paste0(round(ytd_performance, 1), "%"),
    rsi             = round(rsi, 1),
    beta            = round(beta, 2),
    net_margin      = paste0(round(net_margin, 1), "%")
  ) %>%
  rename(
    Ticker     = ticker,
    `YTD %`    = ytd_performance,
    RSI        = rsi,
    Beta       = beta,
    `Net Margin` = net_margin
  ) %>%
  kable(caption = "Top 10 and Bottom 10 YTD Performers — RSI at time of data capture") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped","hover","condensed"),
                position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = COL_PURPLE) %>%
  row_spec(1:10, background = "#d4edda") %>%
  row_spec(11:20, background = "#fde8e8")
```

> **Insight:** The RSI pattern could not be clearer. Every one of the Top 10 performers had RSI in the **43-67 range** (bullish to near-overbought) — stocks already in positive momentum continued gaining. Every one of the Bottom 10 had RSI in the **14-39 range** (oversold to bearish) — stocks already losing momentum continued declining. Critically, the Bottom 10 includes **SIVB (RSI 14.65), SBNY (RSI 25.20), and FRC (RSI 18.52)** — the three banks that failed in March-May 2023. Their catastrophically low RSI was the market's signal that something was deeply wrong, even if traditional financial ratios showed them as "profitable" companies. This is the strongest visual evidence that momentum captures information that fundamentals alone cannot.

## 5.1 Standardized Coefficients

```{r feature_importance_table}
importance_df <- tidy(model_c) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    abs_coef  = abs(estimate),
    Direction = ifelse(estimate > 0, "Positive", "Negative"),
    Significant = ifelse(p.value < 0.05, "Yes", "No")
  ) %>%
  arrange(desc(abs_coef)) %>%
  select(term, estimate, abs_coef, Direction, Significant, p.value) %>%
  mutate(across(c(estimate, abs_coef, p.value), round, 4)) %>%
  rename(
    Variable      = term,
    Coefficient   = estimate,
    `|Coefficient|` = abs_coef,
    `p-value`     = p.value
  )

importance_df %>%
  kable(caption = "Standardized Regression Coefficients — Full Model (Model C)") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped","hover","condensed"),
                position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = COL_PURPLE) %>%
  row_spec(which(importance_df$Significant == "Yes"), bold = TRUE)
```

## 5.2 Feature Importance Chart

```{r feature_importance_chart}
importance_df %>%
  ggplot(aes(x = reorder(Variable, `|Coefficient|`),
             y = Coefficient,
             fill = Direction)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, color = COL_NAVY, linewidth = 0.5) +
  coord_flip() +
  scale_fill_manual(values = c("Positive" = COL_TEAL, "Negative" = COL_RED)) +
  scale_y_continuous(labels = number_format(accuracy = 0.1)) +
  labs(
    title    = "Feature Importance — Standardized Regression Coefficients",
    subtitle = "Predicting YTD Performance | Full Model (Model C)",
    x        = NULL,
    y        = "Standardized Coefficient",
    fill     = "Direction",
    caption  = "Teal = positive predictor of YTD performance | Red = negative predictor | Variables standardized to allow comparison"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 16, color = COL_NAVY),
    plot.subtitle = element_text(color = "gray40"),
    plot.caption  = element_text(color = "gray40"),
    panel.grid.minor  = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position = "bottom"
  )
```

> **Insight:** RSI dominates feature importance by a wide margin — its standardized coefficient (~8.2) is more than twice the next predictor (Beta ~3.7). Among fundamental variables, **Dividend Yield is the most significant negative predictor** — high-yield stocks (energy, banks, utilities) consistently underperformed growth stocks during this period, reflecting the sector rotation from value to growth as interest rate expectations shifted. Analyst recommendation shows a positive coefficient — stocks with stronger buy ratings slightly outperformed, suggesting analyst sentiment has marginal predictive value on top of momentum.

---

# 6. Model Validation

## 6.1 Predicted vs Actual (Test Set)

```{r predicted_vs_actual}
test_results <- data.frame(
  ticker    = test_data$ticker,
  actual    = test_data$ytd_performance,
  predicted = pred_c
) %>% drop_na()

test_r2   <- round(cor(test_results$actual, test_results$predicted)^2, 3)
test_rmse <- round(rmse_fn(test_results$actual, test_results$predicted), 2)

ggplot(test_results, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.55, color = COL_BLUE, size = 2.5) +
  geom_abline(intercept = 0, slope = 1,
              color = COL_RED, linewidth = 1.2, linetype = "dashed") +
  geom_smooth(method = "lm", color = COL_TEAL, se = TRUE, linewidth = 1) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title    = "Predicted vs Actual YTD Performance — Test Set",
    subtitle = paste0("Test R² = ", test_r2, " | RMSE = ", test_rmse, "% | n = ", nrow(test_results)),
    x        = "Actual YTD Performance (%)",
    y        = "Predicted YTD Performance (%)",
    caption  = "Red dashed = perfect prediction | Teal = model fit | Points = individual companies"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 16, color = COL_NAVY),
    plot.subtitle = element_text(color = "gray40"),
    plot.caption  = element_text(color = "gray40"),
    panel.grid.minor = element_blank()
  )
```

> **Insight:** The model explains **`r test_r2*100`%** of YTD performance variation on the held-out test set, with an average prediction error of **±`r test_rmse` percentage points**. The model performs well across the middle range (-10% to +20%) but underestimates the extreme outperformers (NVDA +64%, META +53%) — which is expected, as these moves were driven by the AI narrative emerging in early 2023, a non-quantifiable factor not captured in financial ratios.

## 6.2 Diagnostic Plots

```{r diagnostics, fig.width=12, fig.height=8}
par(mfrow = c(2, 2))
plot(model_c,
     col  = COL_BLUE,
     pch  = 16,
     cex  = 0.6,
     which = 1:4)
par(mfrow = c(1, 1))
```

> **Insight:** Residuals vs Fitted shows no strong non-linear pattern — the linear model is a reasonable approximation. The Q-Q plot shows slight heavy tails, expected given extreme performers like NVDA (+64%). Scale-Location confirms roughly constant variance. No single high-leverage outlier dominates the results.

## 6.3 RSI vs YTD — The Core Relationship

```{r rsi_scatter}
ggplot(df_clean, aes(x = rsi, y = ytd_performance)) +
  geom_point(alpha = 0.45, color = COL_BLUE, size = 2) +
  geom_smooth(method = "lm", color = COL_RED, se = TRUE, linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 70, linetype = "dotted",
             color = COL_RED, alpha = 0.6, linewidth = 1) +
  geom_vline(xintercept = 30, linetype = "dotted",
             color = COL_TEAL, alpha = 0.6, linewidth = 1) +
  annotate("text", x = 73, y = 55, label = "Overbought",
           color = COL_RED, fontface = "bold", size = 3.5, hjust = 0) +
  annotate("text", x = 18, y = 55, label = "Oversold",
           color = COL_TEAL, fontface = "bold", size = 3.5, hjust = 0) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title    = "RSI vs YTD Performance — The Dominant Relationship",
    subtitle = paste0("r = 0.69 | Simple R² = ",
                      round(summary(lm(ytd_performance ~ rsi, data = df_clean))$r.squared, 3)),
    x        = "RSI (Relative Strength Index)",
    y        = "YTD Performance (%)",
    caption  = "Red = overbought zone (RSI > 70) | Teal = oversold zone (RSI < 30)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 16, color = COL_NAVY),
    plot.subtitle = element_text(color = "gray40"),
    plot.caption  = element_text(color = "gray40"),
    panel.grid.minor = element_blank()
  )
```

> **Insight:** The RSI-YTD relationship is striking in its linearity — stocks with RSI already above 50 (in bullish momentum) overwhelmingly continued outperforming, while those below 30 (oversold) had the worst YTD returns. This is the "momentum begets momentum" phenomenon — price strength attracts institutional buying, which drives further price strength. This is one of the most replicated findings in financial economics.

## 6.4 Dividend Yield vs YTD — The Strongest Negative Predictor

```{r dividend_scatter}
ggplot(df_clean %>% filter(dividend_yield <= 10),
       aes(x = dividend_yield, y = ytd_performance)) +
  geom_point(alpha = 0.45, color = COL_ORANGE, size = 2) +
  geom_smooth(method = "lm", color = COL_RED, se = TRUE, linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = mean(df_clean$dividend_yield, na.rm = TRUE),
             linetype = "dotted", color = COL_NAVY, alpha = 0.7, linewidth = 1) +
  annotate("text",
           x = mean(df_clean$dividend_yield, na.rm = TRUE) + 0.3,
           y = 58,
           label = paste0("Mean yield: ",
                          round(mean(df_clean$dividend_yield, na.rm=TRUE),1), "%"),
           color = COL_NAVY, fontface = "bold", size = 3.5, hjust = 0) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(
    title    = "Dividend Yield vs YTD Performance — The Negative Momentum Trap",
    subtitle = paste0("r = -0.30 | High-yield stocks significantly underperformed in early 2023"),
    x        = "Dividend Yield (%)",
    y        = "YTD Performance (%)",
    caption  = "Companies with dividend yield > 10% excluded as outliers | Navy line = mean yield"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 16, color = COL_NAVY),
    plot.subtitle = element_text(color = "gray40"),
    plot.caption  = element_text(color = "gray40"),
    panel.grid.minor = element_blank()
  )
```

> **Insight:** This chart tells the sector rotation story of early 2023 in one visual. High-dividend stocks — energy companies (DVN 8% yield), telecoms (VZ 7%), utilities, banks — were precisely the stocks that lagged as investors rotated from value/income into growth/tech. The negative slope (r = -0.30) means that for every 1 percentage point increase in dividend yield, YTD performance was approximately **`r round(cor(df_clean$dividend_yield, df_clean$ytd_performance, use="complete.obs") * sd(df_clean$ytd_performance, na.rm=TRUE) / sd(df_clean$dividend_yield, na.rm=TRUE), 1)`** percentage points lower on average. The stocks paying the most income to shareholders were precisely those being sold by the market.

## 7.1 Summary of Key Findings

```{r summary_table}
findings <- data.frame(
  Section = c(
    "Data Quality",
    "Profitability",
    "Valuation",
    "Risk",
    "Growth",
    "Dividends",
    "Momentum",
    "Forecasting"
  ),
  Key_Finding = c(
    "PEG ratio 23.2% missing; 3 failed banks appeared undervalued — quantitative screens cannot detect bank run risk",
    "NDX companies average 57.94% gross margin; 7 triple-index companies average 71.57% ROE — quality premium is real",
    "Energy sector trades at PEG 0.14-0.44 (structurally discounted for ESG risk); NVDA at PEG 6.55 priced for AI dominance",
    "15 companies (3%) pass the 'Financial Fortress' screen; DJIA companies most leveraged (D/E 2.97) but most defensive (Beta 0.87)",
    "HES has 78.1% long-term growth estimate at PEG 0.26 — energy ESG discount creates the widest growth/valuation gap",
    "DVN pays 8.02% yield but it is variable (oil-price-linked); IBM's 362.9% payout ratio is not sustainable",
    "Momentum (RSI) correlates 0.69 with YTD performance — the dominant short-term driver",
    "Momentum Model R² = 0.56 vs Fundamentals Model R² = 0.16 — short-run performance is momentum-driven, not fundamental"
  )
)

findings %>%
  kable(caption = "Key Findings by Section",
        col.names = c("Section", "Key Finding")) %>%
  kable_styling(full_width = TRUE,
                bootstrap_options = c("striped","hover","condensed"),
                position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = COL_PURPLE)
```

## 7.2 Business Recommendations

Based on the MySQL analysis (17 queries) and R regression model, five actionable recommendations emerge:

**1. Use Momentum as the Primary Short-Term Filter**
RSI alone explains 56% of short-term returns. Any stock screen for near-term performance should incorporate RSI before fundamental metrics. Stocks in the 50-70 RSI zone (bullish but not overbought) historically offer the best risk-adjusted entry points.

**2. Recognize the Energy Sector Structural Discount**
Energy companies (HES, MRO, XOM, EOG) have PEG ratios of 0.14-0.44 — the market assigns a permanent ESG/energy-transition discount that is not reflected in current earnings or growth estimates. Long-term investors willing to hold through the energy transition may be overpaid in fundamentals relative to their price.

**3. The 'Financial Fortress' Screen as a Quality Baseline**
Only 15 of 496 companies simultaneously pass all four quality criteria (D/E < 0.5, Current Ratio > 2, Net Margin > 10%, Beta < 1.2). VRTX, REGN, BIIB, GOOGL, and Mastercard stand out as genuinely high-quality businesses without financial engineering inflating their metrics.

**4. Use PEG Rather Than P/E for Growth Stocks**
P/E alone is misleading for companies with high growth expectations. Airlines (UAL PEG 0.44, DAL PEG 0.52) appear cheap by PEG despite high P/E — because their post-COVID earnings recovery was still in early stages. PEG below 1.0 combined with analyst "Buy" ratings is a more reliable combined signal.

**5. Data Timestamp Risk**
The most important lesson: SVB, Signature Bank, and First Republic all appeared financially healthy in this dataset — they failed months later. No financial ratio captures bank run risk, deposit concentration risk, or duration mismatch in bond portfolios. Quantitative analysis must always be supplemented with qualitative assessment of business model risks.

## 7.3 Methodological Limitations

```{r limitations}
limitations <- data.frame(
  Limitation = c(
    "Single time snapshot",
    "REIT metrics distortion",
    "Debt-inflated ROE",
    "Survivorship bias",
    "No sector controls",
    "RSI endogeneity"
  ),
  Description = c(
    "Dataset captures a single point in early 2023. Results reflect bear market recovery dynamics and may not generalize to other market regimes",
    "REITs show 100%+ gross margins and 200%+ payout ratios that are structurally different from regular companies — these metrics are not comparable across sectors",
    "Companies with tiny/negative equity show extreme ROE (APA 924.9%, MSI -518.7%) that reflects leverage, not operational excellence — these were removed from the regression",
    "Dataset only includes companies currently in the S&P 500, not those that were delisted or failed — introducing survivorship bias",
    "Sector membership is a major omitted variable — energy, banks, and tech have very different business cycles that explain much of the YTD variation",
    "RSI is derived from past price performance, which already contains YTD information — the correlation partly reflects circular measurement rather than pure causation"
  )
)

limitations %>%
  kable(caption = "Methodological Limitations") %>%
  kable_styling(full_width = TRUE,
                bootstrap_options = c("striped","hover","condensed"),
                position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = COL_PURPLE)
```

---

# 8. Future Work

- **Sector-controlled analysis:** Run separate regressions within each GICS sector to isolate within-sector predictors, removing the sector effect from coefficient estimates
- **Time-series extension:** Apply the same pipeline quarterly across 2020-2024 to test whether momentum dominance is consistent across different market regimes (bull, bear, recovery)
- **Alternative ML models:** Compare linear regression to Random Forest and XGBoost to capture non-linear interactions (e.g. the interaction between RSI and PE ratio for momentum + value stocks)
- **Portfolio simulation:** Construct a paper portfolio based on the composite scorecard (Query 8.1) and backtest it against the S&P 500 index return
- **Live data integration:** Connect to Yahoo Finance API via `quantmod` for real-time metric updates rather than a static snapshot

---

*Report generated using R `r R.version$major`.`r R.version$minor` | Data source: MySQL sp500_project.companies (Query 9.1 export) | Analysis tools: tidyverse, caret, corrplot, broom*
